Optimality and Heuristics for Finite Element Integration

### 2015-10-29 TOMS_Paper/assembly.tex
Fixes to shar el. Also, trading opt for analysis cost

To synthesize optimal implementations as in Definition~\ref{def:mln-optimality} we need:
\begin{itemize}
\item a characterization of the space of legal transformations impacting the operation count
\item a cost model to select the optimal point in the transformation space
\end{itemize}
In the following Sections~\ref{sec:sharing-elimination} and~\ref{sec:pre-evaluation}, we focus on the construction of the transformation space.

@@

In this section, we study an optimal transformation to perform \textit{sharing elimination}; that is, a transformation that minimizes the operation count of a loop nest by applying factorization, code motion and (if necessary) common sub-expression elimination. In general, transformations of this sort require solving large combinatorial problems; for instance, to evaluate the impact of all possible factorization strategies. However, we show that by restricting ourselves to the class of loops and expressions of Definition~\ref{def:fem-loopnest}, the search space can drastically be pruned. In the following, we assume outer product loop nests of arity $n=2$ (i.e., a bilinear form), although the same reasoning stands for the case $n < 2$.

@@

\begin{enumerate}
\item Assume $\alpha \neq \beta$; that is, test and trial functions belong to different function spaces. By factorizing, in sequence, any $a_{ij} \in \alpha_{eij}$ and then any $b_{ik} \in \beta_{eik}$ (or vice-versa), sub-expressions independent of one of the outer product loops are exposed. This systematically exposes all code motion opportunities to minimize the operation count in the innermost loop.
\item In the special case $\alpha = \beta$, common sub-expressions arise over different iteration spaces ($ij$ and $ik$). Hoisting and capturing these sub-expressions may make the operation count of outer loops lower than in point 1), but the gain (if any) will be negligible (note: the operation count in the innermost loop does not vary).
\item Since loop trip counts are never lower than 3 (P1 Lagrange elements on triangles), it is never the case that applying points 1) or 2) increases the operation count (it might have been the case, otherwise, in a scenario in which loops degenerate, i.e. with iteration spaces of size 1).
\item Factorizing any of the terms in $\sigma_{ei}$ (e.g., a coefficient, or a coefficient derivative along a certain spatial dimension) can have multiple implications. We distinguish two cases:
\begin{enumerate}
\item Factorization occurs in place of points 1) or 2). This either (i) inhibits (decrease) code motion by producing non-factorizable, non-hoistable sub-expressions or (ii) simply reduces to cases 1) or 2) (for example, in the case of forms with scalar coefficients).
\item Factorization occurs after points 1) or 2); for example, we have $a_{ij}(b_{ik}(\sigma'_{ei} + \sigma''_{ei}))$, with $\sigma_{ei} = \sigma'_{ei} + \sigma''_{ei}$. Splitting $L_e$-dependent from $L_i$-dependent sub-expressions (through suitable re-factorization) and applying code motion naturally tends to operation count minimization. If the loop trip count is extremely low, however, this strategy may be sub-optimal. For example, consider the following sub-expression, $(a v_i + b w_i) (c v_i + d w_i)$. It is true that factorizing $v_i$ and $w_i$ would expose $L_i$-independent sub-expressions, but this would actually result in increased operation count if, for instance, the size of $L_i$ were 1.
\end{enumerate}
\end{enumerate}

Based on these observations, in Algorithm~\ref{algo:opt} we explain the sequence of transformations for potentially optimal sharing elimination. The algorithm does not take into account the concerns raised in points 2) and 4b). To address them, the simplest solution consists of comparing the operation counts by generating the corresponding code variants. We will further comment on this in Section~\ref{sec:obsheu}.

\begin{Algo}[Construction of the potential optimum]
\label{algo:opt}
\normalfont
Consider a finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$. The algorithm starts with reducing the input expression to the normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$.

The sub-expressions $\alpha_{eij}^{w} \beta_{eik}^{w}$ are ``flattened'' (by multiplication expansion) such that all $L_j$- and $L_k$-dependent terms lie on the same level of the expression tree. We factorize, in sequence, $L_j$- ($L_k$-) and $L_k$- ($L_j$-) dependent terms (it can easily be demonstrated, by contradiction, that one should start from the loop with the least number of unique terms in order to obtain the smallest operation count). Thanks to linearity, each factored product only has one symbol depending on $L_j$ ($L_k$), and this symbol is now unique in the expression. The other sub-expression, independent of $L_j$ ($L_k$), is, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided. Note that multilinearity ensures deterministic factorization, while perfectness ensures correctness of hoisting. Finally, we factorize $L_i$-dependent terms such that $L_e$-dependent sub-expressions are hoisted.
\end{Algo}

@@

If, on the other hand, $\iota < I$, a further step is necessary to estimate $\theta_{pre}$. In particular, we need to calculate the number of terms $\rho$ such that $\theta_{pre} = \rho \cdot \iota$. Consider again Figure~\ref{code:increase_factor}. In the case of the mass matrix, the body of $L_k$ is simply characterized by the dot product of test and trial functions, \texttt{B[j]*B[k]}, so trivially $\rho = 1$. In general, $\rho$ varies with the discretization and the differential operators used in the form. For example, in the case of the bilinear form originating from a standard bi-dimensional Poisson equation, the reader could verify that after a suitable factorization we would have $\rho = 3$. There are several ways of determining $\rho$. The fastest would be to extract it from high-level analysis of form and discretization; for convenience, in our implementation we have algorithms that, based on analysis of the expression tree, project the output of monomial expansion and factorization, which in turn gives us $\rho$.

\subsection{Trading optimality for analysis cost}
\label{sec:obsheu}
\begin{itemize}
\item SPEED UP THETA: We can easily predict $\theta_{se}$ thanks to our key property, linearity. This was explained in Proposition~\ref{prop:optimal-mln}. A simple expression tree traversal suffices to obtain the cost of a  sharing-free outer product loop nest, namely $\Lambda_{ops}^{ns}$. Assuming $I$ to be the size of $L_i$, we have that $\theta_{se} = \Lambda_{ops}^{ns} \cdot I$.
\item SPEED UP ALGO: most of the time, only a single bipartition...
\item SPEED UP SHARING EL: stick to opt
\end{itemize}

>>>>

To synthesize optimal implementations as in Definition~\ref{def:mln-optimality} we need:
\begin{itemize}
\item a characterization of the space of legal transformations impacting the operation count
\item a cost model to select the optimal point in the transformation space
\end{itemize}
In the following Sections~\ref{sec:sharing-elimination} and~\ref{sec:pre-evaluation}, we focus on the construction of the transformation space. We consider outer product loop nests of arity $n=2$ (i.e., bilinear forms), although a similar (simpler) reasoning straightforwardly applies to the case $n < 2$.

@@

In this section, we study optimal \textit{sharing elimination}; that is, a transformation that minimizes a loop nest operation count by applying factorization, code motion and (if necessary) common sub-expression elimination. In general, transformations of this sort require solving large combinatorial problems; for instance, to evaluate the impact of all possible factorization strategies. However, we explain that by restricting ourselves to the class of loops and expressions in Definition~\ref{def:fem-loopnest}, the search space becomes very small.

@@

\begin{enumerate}
\item Assume $\alpha \neq \beta$; that is, test and trial functions belong to different function spaces. By factorizing, in sequence, any $a_{ij} \in \alpha_{eij}$ and then any $b_{ik} \in \beta_{eik}$ (or vice-versa), sub-expressions independent of one of the outer product loops are exposed. This systematically exposes all code motion opportunities to minimize the operation count in the innermost loop.
\item In the special case $\alpha = \beta$, common sub-expressions arise over different iteration spaces ($ij$ and $ik$). Compared to point 1), by hoisting and capturing these sub-expressions the reduction loop operation count becomes smaller, although the gain tends to be negligible (the operation count in the innermost loop is not affected).
\item If we neglect the pathological cases in which the loops degenerate (i.e., when the trip count becomes 1, which is the case of P0 discontinuous elements), factorizing and hoisting as in points 1) or 2) systematically reduces the operation count. The reduction is proportional to the outer product loop nest extent.
\item Factorizing any of the terms in $\sigma_{ei}$ (e.g., a coefficient, or a coefficient derivative along a certain spatial dimension) also impacts the operation count. We distinguish two cases:
\begin{enumerate}
\item Test/trial functions are yet to be factorized. This, in general, inhibits (decreases) code motion by producing non-factorizable, non-hoistable sub-expressions. In the best case (e.g., there is a single scalar coefficient pre-multiplying the form) this simply reduces to cases 1) or 2).
\item Within $\sigma_{ei}$ itself; for example, after applying point 1) we have $a_{ij}(b_{ik}(\sigma'_{ei} + \sigma''_{ei}))$, with $\sigma_{ei} = \sigma'_{ei} + \sigma''_{ei}$. Splitting $L_e$-dependent from $L_i$-dependent sub-expressions (through suitable re-factorization) and applying code motion naturally tends to operation count minimization. Similarly to point 3), however, with very low trip counts this strategy may be sub-optimal. For example, consider the following sub-expression, $(a v_i + b w_i) (c v_i + d w_i)$, which is excerpted from a multiplication of the inverse Jacobian matrix for the change of coordinates with the derivative of a vector-valued coefficient. It is true that factorizing $v_i$ and $w_i$ exposes $L_i$-independent sub-expressions, but this will actually increase the operation count when the size of $L_i$ is 1.
\end{enumerate}
\end{enumerate}

Algorithm~\ref{algo:opt} turns these observations into a sequence of steps leading to optimal sharing elimination.

\begin{Algo}[Sharing elimination]
\label{algo:opt}
\normalfont
Consider a finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$. The algorithm starts with reducing the input expression to the normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$. The sub-expressions $\alpha_{eij}^{w} \beta_{eik}^{w}$ are then ``flattened'' (by multiplication expansion) such that all $L_j$- and $L_k$-dependent terms lie on the same level of the expression tree. We factorize, in sequence, $L_j$- ($L_k$-) and $L_k$- ($L_j$-) dependent terms. The choice concerning what factorize first, which impacts the resulting operation count, depends on two factors: the number of unique terms in each loop, $n$, and the trip counts, $tc$. It is fairly simple to see (and easily demonstrable by contradiction) that by starting from the loop such that $n tc$ is smallest leads to the maximal operation count reduction. Assume then we start with $L_j$. Thanks to linearity, each factored product only has one symbol depending on $L_j$ ($L_k$), and this symbol is now unique in the expression. The other sub-expression, independent of $L_j$ ($L_k$), is, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided. Note that multilinearity ensures deterministic factorization, while perfectness ensures correctness of hoisting. Finally, we factorize $L_i$-dependent terms such that $L_e$-dependent sub-expressions are hoisted, checking that this actually results in reduced operation count (see point 4b) above).
\end{Algo}

@@

If, on the other hand, $\iota < I$, a further step is necessary to estimate $\theta_{pre}$. In particular, we need to calculate the number of terms $\rho$ such that $\theta_{pre} = \rho \cdot \iota$. Consider again Figure~\ref{code:increase_factor}. In the case of the mass matrix, the body of $L_k$ is simply characterized by the dot product of test and trial functions, \texttt{B[j]*B[k]}, so trivially $\rho = 1$. In general, $\rho$ varies with the discretization and the differential operators used in the form. For example, in the case of the bilinear form originating from a standard bi-dimensional Poisson equation, the reader could verify that after a suitable factorization we would have $\rho = 3$. To determine $\rho$, we simply apply Algorithm~\ref{algo:pre-evaluation} until the evaluation of the reduction (obviously, excluded).

\subsection{Trading optimality for analysis cost}
Code generation time is fundamental for productivity. In this section, we discuss how to reduce it by relaxing the constraints on optimality.
\label{sec:obsheu}
\begin{itemize}
\item \textit{Sharing elimination.} In Section~\ref{sec:sharing-elimination}, we explained that very small loop trip counts render certain factorizations sub-optimal. To reach optimality, Algorithm~\ref{algo:opt} needs to backtrack if it detects an increase in operation count after applying sharing elimination at a given loop nest level. Since this situation is particularly rare, we can relieve the algorithm by avoiding these checks.
\item \textit{Transformation algorithm.} The transformation algorithm (Figure~\ref{code:intuition}) evaluates all possible ways of scheduling monomials candidate to pre-evaluation. In extremely complex forms, this could increase the analysis cost. Despite we have never encountered such a situation, a possibility is to disable these checks and to apply the transformation to all candidates. In the worst case scenario, this leads to a local optimum.
\item \textit{Construction of $\theta$.} Instead of determining $\theta_{se}$ and $\rho$ (for $\theta_{pre}$) by manipulating the expression tree, we can approximate these values exploiting the key property of outer product loop nests, linearity. The number of unique terms in a multilinear loop nest can be used to calculate an upper bound on the cost of the innermost loop, where the bulk of the computation naturally tends to be performed. For example, in Figure~\ref{code:multi_loopnest_a} the fact that only a single term depends on $L_j$, $b_j$, suggests that in the innermost loop no more than one multiplication will be performed once sharing is eliminated. Therefore, an approximation for $\theta_{se}$ is, for example, $\theta_{se} = n_{iterations} n_{unique}$, where $n_{iterations}$ is the total number of iterations, and $n_{unique}$ is the smallest number of unique terms along one the linear loops.
\end{itemize}

In our code generation system, described in the next section, we make use of all of these heuristics.
